---
title: "Lab 10: Review Lab: Hypothesis Testing and Confidence Intervals"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This lab reviews core concepts in hypothesis testing using a real dataset. The goal is not only to perform statistical tests, but also to understand when to use them, how to interpret results, and how they connect to confidence intervals.

We will cover:
1. Paired t-tests  
2. Two-sample t-tests  
3. One-sample tests  
4. Proportion tests  
5. Confidence intervals  
6. Type I and Type II errors  
7. Choosing the correct test  

---

# Load Data

```{r}
lab_data <- read.csv("hypothesis_lab_data.csv")
head(lab_data)
summary(lab_data)
```

---

# Part 0: Understanding the Data

## Question 1

Classify each of the following variables as either categorical or numerical. Briefly justify your answer.

- group  
- pre_score  
- post_score  
- improvement  
- passed  

## Answer

- group: Categorical variable because it represents group membership (Control vs Tutoring).  
- pre_score: Numerical variable because it measures test scores.  
- post_score: Numerical variable because it measures test scores.  
- improvement: Numerical variable because it represents a computed difference in scores.  
- passed: Categorical variable because it indicates a binary outcome (Yes or No).  

---

## Question 2

Provide at least three meaningful statistical questions that can be investigated using this dataset.

## Answer

- Did students improve after the intervention (pre vs post scores)?  
- Did the tutoring group improve more than the control group?  
- Is the average post-test score significantly different from 75?  
- Is the proportion of students who passed greater than 70%?  

---

## Exploratory Data Analysis

```{r}
# Summary by group
aggregate(improvement ~ group, data = lab_data, mean)

# Boxplot
boxplot(improvement ~ group, data = lab_data, col = c("lightblue", "lightgreen"),
        main = "Improvement by Group")

# Histogram
hist(lab_data$improvement, main = "Distribution of Improvement", col = "lightgray")
```

---

# Part 1: Paired t-test

## Research Question

Do students show a statistically significant improvement from pre-test to post-test scores?

---

## Question 1

Clearly state the null and alternative hypotheses.

## Answer

H0: The mean difference between post-test and pre-test scores is equal to 0.  
HA: The mean difference between post-test and pre-test scores is greater than 0.  

---

## Question 2

Explain why a paired t-test is appropriate in this setting.

## Answer

A paired t-test is appropriate because the pre-test and post-test scores are measured on the same individuals. This creates dependent observations, and the test accounts for within-subject variation.

---

## Question 3

Conduct the paired t-test.

```{r}
paired_test <- t.test(lab_data$post_score, lab_data$pre_score,
                     paired = TRUE, alternative = "greater")
paired_test
```

---

## Question 4

Interpret the p-value in the context of this problem.

## Answer

If the p-value is less than 0.05, we reject the null hypothesis and conclude that there is statistically significant evidence that students improved after the intervention.

---

## Question 5

Compute and interpret the confidence interval for the mean difference.

```{r}
paired_test$conf.int
```

## Answer

If the confidence interval lies entirely above 0, this indicates that the average improvement is positive, reinforcing the conclusion that students improved.

---

## Question 6

Explain what a Type I error and a Type II error would mean in this context.

## Answer

A Type I error would occur if we conclude that students improved when, in reality, there was no improvement.  
A Type II error would occur if we fail to detect an actual improvement in student performance.

---

# Part 2: Two-Sample t-test

## Research Question

Does the tutoring group experience greater improvement than the control group?

---

## Question 1

State the null and alternative hypotheses.

## Answer

H0: The mean improvement in the tutoring group is equal to the mean improvement in the control group.  
HA: The mean improvement in the tutoring group is greater than that of the control group.  

---

## Question 2

Conduct the two-sample t-test.

```{r}
two_sample_test <- t.test(improvement ~ group, data = lab_data,
                         alternative = "greater")
two_sample_test
```

---

## Question 3

Interpret the results in context.

## Answer

If the p-value is less than 0.05, we conclude that the tutoring group shows significantly greater improvement compared to the control group.

---

# Part 3: One-Sample Mean Test

## Research Question

Is the average post-test score significantly different from 75?

---

## Question 1

State the hypotheses.

## Answer

H0: The mean post-test score is equal to 75.  
HA: The mean post-test score is not equal to 75.  

---

## Question 2

Conduct the test.

```{r}
one_sample_test <- t.test(lab_data$post_score, mu = 75)
one_sample_test
```

---

## Question 3

Interpret the results.

## Answer

If the p-value is less than 0.05, we conclude that the mean post-test score is significantly different from 75.

---

# Part 4: Proportion Test

## Research Question

Is the proportion of students who passed greater than 70%?

---

## Question 1

State the hypotheses.

## Answer

H0: The proportion of students who passed is equal to 0.70.  
HA: The proportion of students who passed is greater than 0.70.  

---

## Question 2

Conduct the proportion test.

```{r}
prop_test <- prop.test(sum(lab_data$passed == "Yes"),
                       nrow(lab_data),
                       p = 0.7,
                       alternative = "greater")
prop_test
```

---

## Question 3

Interpret the results.

## Answer

If the p-value is less than 0.05, we conclude that the proportion of students who passed is significantly greater than 70%.

---

# Part 5: Confidence Intervals

## Question

Explain the relationship between hypothesis testing and confidence intervals.

## Answer

Confidence intervals provide a range of plausible values for a parameter. If the null hypothesis value lies outside the confidence interval, we reject the null hypothesis. If it lies within the interval, we fail to reject the null hypothesis.

---

# Part 6: Choosing the Correct Test

## Question

Match each scenario with the appropriate statistical test.

1. Measurements taken before and after on the same individuals  
2. Comparing two independent groups  
3. Comparing a sample mean to a known value  
4. Testing a proportion  

## Answer

1. Paired t-test  
2. Two-sample t-test  
3. One-sample t-test  
4. Proportion test  

---

# Part 7: Assumptions

## Question

List and explain the key assumptions required for t-tests.

## Answer

- Independence: Observations must be independent of each other.  
- Normality: The data should be approximately normally distributed, or the sample size should be sufficiently large.  
- Equal variance (for two-sample tests): The variability in both groups should be similar, depending on the test used.  

---

# Part 8: Effect Size vs Statistical Significance

## Question

Explain the difference between statistical significance and practical significance.

## Answer

Statistical significance indicates whether an observed effect is unlikely to be due to chance. Practical significance considers whether the magnitude of the effect is meaningful in a real-world context. A result can be statistically significant but still have little practical importance.

---

# Final Reflection

## Question

What is the most important takeaway from hypothesis testing?

## Answer

Hypothesis testing provides a structured framework for making decisions based on data, but results must always be interpreted carefully in context, considering assumptions, effect sizes, and real-world implications.