---
title: "Lab 9: Causal Inference with Difference-in-Differences"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---
# Introduction

In this lab, we will study a classic causal inference question:

**Does raising the minimum wage reduce employment?**

We will use a fast-food restaurant dataset modeled after the well-known Card and Krueger setting:
- **New Jersey (NJ)** increased its minimum wage
- **Pennsylvania (PA)** did not

This gives us a natural treatment and control setting:

- **Treatment group**: New Jersey restaurants
- **Control group**: Pennsylvania restaurants
- **Before period**: before the policy change
- **After period**: after the policy change

Our goals are:

1. Understand the logic of **Difference-in-Differences (DiD)**
2. Compare employment before and after the policy
3. Compare wages before and after the policy
4. Estimate the causal effect of the minimum wage increase on:
   - **employment**
   - **wages**

---

# Part 1: Load Packages and Data

## Why are we doing this?

Before we can analyze the data, we need to:
- load the packages we will use
- read the CSV file
- inspect the dataset
- make sure variables are stored in the correct format

Because your `fastfood.csv` file is in the **same folder** as the `.qmd` file, we can read it directly with `read.csv("fastfood.csv")`.

```{r}
# Set CRAN mirror so Quarto does not stop and ask for one during rendering
options(repos = c(CRAN = "https://cloud.r-project.org"))

# List the packages we need for this lab
required_packages <- c("tidyverse", "broom")

# Check which packages are not installed yet
new_packages <- required_packages[
  !(required_packages %in% installed.packages()[, "Package"])
]

# Install any missing packages
if(length(new_packages)){
  install.packages(new_packages)
}

# Load all required packages
lapply(required_packages, library, character.only = TRUE)

# Read the CSV file from the same folder as this .qmd file
fastfood <- read.csv("fastfood.csv")

# Show the structure of the dataset
str(fastfood)

# Show the first 6 rows
head(fastfood)


```

# Part 2: Understand the Variables
## Why are we doing this?

Before running any model, we need to understand:

- what each variable means
- which variables are categorical
- which variables are numeric
- which variables we will use in the DiD setup

```{r}
# Convert state, time, and chain into categorical variables (factors)
fastfood$state <- factor(fastfood$state)
fastfood$time  <- factor(fastfood$time)
fastfood$chain <- factor(fastfood$chain)

# Check the structure again after converting variables
str(fastfood)

# Show summary statistics for all variables
summary(fastfood)
```

## Key Variables for DiD
store_id: unique restaurant identifier
state: NJ or PA
time: Before or After
fte: employment measured as full-time equivalent workers
wage: hourly wage
chain: restaurant chain
treat: indicator for treatment group
post: indicator for after period
did: interaction term

## Critical Question:
Why do we need both a treatment group and a control group? Why is it not enough to only look at New Jersey before and after the policy?

If we only looked at New Jersey before and after the policy, we would not know whether any employment change was caused by the minimum wage increase or by something else happening over time. For example, employment could change because of seasonal demand, local economic conditions, or industry trends. The control group gives us a benchmark for what might have happened even without the policy.

# Check for missing values
Missing values can cause:
errors in regressions
dropped observations
misleading results

So we should always check for them early.
```{r}
# Count missing values in each column
colSums(is.na(fastfood))
```
## Critical Question:
Why is it important to check for missing values before running a regression?

If important variables have missing values, R may silently remove some rows from the analysis. This changes the sample size and can affect results. If the missing data are systematic rather than random, the regression results may also become biased.

# Create the DiD variables
Even though the dataset already includes treat, post, and did, it is good practice to recreate them ourselves. This helps us understand how the design works.

In a DiD setup:
treat = 1 if the restaurant is in NJ, otherwise 0
post = 1 if the observation is after the policy, otherwise 0
did = treat * post
```{r}
# Recreate treatment indicator: 1 for NJ, 0 for PA
fastfood$treat_new <- ifelse(fastfood$state == "NJ", 1, 0)

# Recreate post indicator: 1 for After, 0 for Before
fastfood$post_new <- ifelse(fastfood$time == "After", 1, 0)

# Create the DiD interaction term
fastfood$did_new <- fastfood$treat_new * fastfood$post_new

# Inspect the first 10 rows of the key variables
fastfood %>%
  select(store_id, state, time, treat_new, post_new, did_new) %>%
  head(10)
```

# Critical Question
In this lab, which observations have did_new = 1?

Only observations that are both:
in New Jersey
in the After period
These are the treated observations after the policy change.

# Part 3: Descriptive Analysis
## Why are we doing this?
Before any regression, we want to build intuition by looking at:
average employment by state and time
average wages by state and time

This helps us understand the raw patterns in the data.
```{r}
# Compute average employment and wage by state and time
group_summary <- fastfood %>%
  group_by(state, time) %>%
  summarise(
    mean_fte = mean(fte),
    mean_wage = mean(wage),
    n = n(),
    .groups = "drop"
  )

# Display the summary table
group_summary
```
## Critical Question
If average wages increase much more in NJ than in PA, what does that suggest?

It suggests that the policy likely affected wages in NJ, since NJ was the treated state. This is exactly what we would expect if the minimum wage increase was binding and had a real effect on pay.

# Part 4: Visualize Employment and Wages
## Why are we doing this?
Plots make the before-and-after patterns easier to see.

We will create:
- an employment plot
- a wage plot
```{r}
# Plot average employment by state and time
fastfood %>%
  group_by(state, time) %>%
  summarise(mean_fte = mean(fte), .groups = "drop") %>%
  ggplot(aes(x = time, y = mean_fte, group = state, linetype = state)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Employment by State and Time",
    x = "Time Period",
    y = "Average FTE Employment"
  )
```

```{r}
# Plot average wage by state and time
fastfood %>%
  group_by(state, time) %>%
  summarise(mean_wage = mean(wage), .groups = "drop") %>%
  ggplot(aes(x = time, y = mean_wage, group = state, linetype = state)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Wage by State and Time",
    x = "Time Period",
    y = "Average Hourly Wage"
  )
```

## Critical Question
In a DiD design, what would we expect the wage plot to look like if the treatment had a strong effect on wages?

We would expect the NJ line to rise noticeably after the policy, while the PA line would remain relatively flat or change much less. That would show a larger change in wages for the treated group.

# Part 5: Manual DiD Calculation for Employment
## Why are we doing this?
Before running a regression, we can calculate the DiD estimate manually using the group means. This helps us understand the logic of the DiD estimator. The formula is:
DiD = (Y_NJ,After - Y_NJ,Before) - (Y_PA,After - Y_PA,Before)

```{r}
# Compute average employment in each group
employment_means <- fastfood %>%
  group_by(state, time) %>%
  summarise(mean_fte = mean(fte), .groups = "drop")

employment_means
```

```{r}
# Extract the four means needed for DiD
nj_before_fte <- employment_means %>%
  filter(state == "NJ", time == "Before") %>%
  pull(mean_fte)

nj_after_fte <- employment_means %>%
  filter(state == "NJ", time == "After") %>%
  pull(mean_fte)

pa_before_fte <- employment_means %>%
  filter(state == "PA", time == "Before") %>%
  pull(mean_fte)

pa_after_fte <- employment_means %>%
  filter(state == "PA", time == "After") %>%
  pull(mean_fte)

# Compute within-state changes
nj_change_fte <- nj_after_fte - nj_before_fte
pa_change_fte <- pa_after_fte - pa_before_fte

# Compute the DiD estimate
did_fte_manual <- nj_change_fte - pa_change_fte

# Print the results
nj_before_fte
nj_after_fte
pa_before_fte
pa_after_fte
nj_change_fte
pa_change_fte
did_fte_manual
```

## Critical Question:
Why do we subtract PA's change from NJ's change?

Because PA represents the untreated comparison group. Its change over time captures general time trends that may have affected employment regardless of the policy. By subtracting PA's change, we isolate the extra change associated with treatment.

# Part 6: Manual DiD Calculation for Wages
```{r}
# Compute average wages in each group
wage_means <- fastfood %>%
  group_by(state, time) %>%
  summarise(mean_wage = mean(wage), .groups = "drop")

wage_means
```

```{r}
# Extract the four means needed for wage DiD
nj_before_wage <- wage_means %>%
  filter(state == "NJ", time == "Before") %>%
  pull(mean_wage)

nj_after_wage <- wage_means %>%
  filter(state == "NJ", time == "After") %>%
  pull(mean_wage)

pa_before_wage <- wage_means %>%
  filter(state == "PA", time == "Before") %>%
  pull(mean_wage)

pa_after_wage <- wage_means %>%
  filter(state == "PA", time == "After") %>%
  pull(mean_wage)

# Compute within-state changes
nj_change_wage <- nj_after_wage - nj_before_wage
pa_change_wage <- pa_after_wage - pa_before_wage

# Compute the DiD estimate for wages
did_wage_manual <- nj_change_wage - pa_change_wage

# Print the results
nj_before_wage
nj_after_wage
pa_before_wage
pa_after_wage
nj_change_wage
pa_change_wage
did_wage_manual
```

## Critical Question:
Suppose wage DiD is clearly positive, but employment DiD is close to zero or positive. What does that suggest?

It suggests that the minimum wage increase raised wages without causing a corresponding decrease in employment. This would go against the simplest version of the standard competitive labor market prediction.

# Part 7: Regression-Based DiD Estimation for Employment
The manual DiD estimate is useful, but regression gives us a standard, formal framework. The formula is:
Y = β0 + β1*treat + β2*post + β3*did + ε

Where:
- Y is the outcome (employment or wages)
- treat is the treatment group indicator
- post is the after period indicator
- did is the interaction term (treat * post)    
```{r}
# Run the DiD regression for employment
model_fte <- lm(fte ~ treat_new + post_new + did_new, data = fastfood)

# Show regression output
summary(model_fte)
```

## Critical Question:
In the employment regression, which coefficient should we focus on if we want the estimated causal effect of the policy?

The coefficient on did_new, because it measures the change in NJ after the policy relative to the change in PA over the same period.

# Part 8: Regression-Based DiD Estimation for Wages
```{r}
# Run the DiD regression for wages
model_wage <- lm(wage ~ treat_new + post_new + did_new, data = fastfood)

# Show regression output
summary(model_wage)
```

## Critical Question:
Why is it useful to estimate the wage effect as well as the employment effect?

Because if the minimum wage policy truly changed labor market conditions, we should first see evidence that wages increased in the treated state. Wage results help verify that the treatment actually affected the intended variable. If wages did not change, then we would not expect to see any employment effect either.   

# Part 9: Tidy Regression Tables
```{r}
# Create a tidy table for the employment model
tidy(model_fte)

# Create a tidy table for the wage model
tidy(model_wage)
```

# Part 10: Add Chain Controls
## Why are we doing this?
Adding controls can help improve the precision of our estimates and account for any differences in chain distribution between states. For example, if NJ has more McDonald's locations and McDonald's pays higher wages, that could confound our estimates. By controlling for chain, we can better isolate the effect of the minimum wage policy.  

```{r}
# Employment model with chain controls
model_fte_controls <- lm(fte ~ treat_new + post_new + did_new + chain, data = fastfood)

# Wage model with chain controls
model_wage_controls <- lm(wage ~ treat_new + post_new + did_new + chain, data = fastfood)

# Show summaries
summary(model_fte_controls)
summary(model_wage_controls)
```

## Critical Question:
Why might controlling for chain be helpful in this setting?

Some chains may systematically pay higher wages or employ more workers than others. If chain composition differs between states, then failing to control for chain could mix treatment effects with chain-specific differences.

# Part 11: Compare Manual and Regression DiD Estimates
```{r}
# Extract the regression DiD coefficients
coef(model_fte)["did_new"]
coef(model_wage)["did_new"]

# Show the manual DiD estimates again
did_fte_manual
did_wage_manual
```

In a simple DiD setup with no additional controls, the manual DiD estimate should equal the regression coefficient on the interaction term.

## Critical Question:
Why do the manual DiD estimate and the regression interaction coefficient match here?

Because the regression is just a formal way of expressing the same group-mean comparison. In the basic two-group, two-period DiD setup, the interaction coefficient reproduces the same difference-in-differences calculation.

# Assumptions behind DiD
- Parallel trends: In the absence of treatment, the average outcomes for treated and control groups would have followed parallel paths over time.
- No spillovers: The treatment in NJ does not affect outcomes in PA.
- Stable composition: The set of restaurants in each state does not change systematically over time in a way that is related to the treatment.  

In this dataset, we only have one before period and one after period, so we cannot directly test parallel trends well. Because with only one pre-treatment observation, we cannot observe an actual trend before treatment. We can compare levels before treatment, but we cannot see whether the two states were moving similarly over time before the policy.

## Critical Question:
If NJ and PA were already on very different employment trends before treatment, how would that affect our DiD estimate?

Then the DiD estimate might incorrectly attribute those pre-existing trend differences to the policy. In that case, the causal interpretation would be weaker.

# Conclusions
## Employment
The employment DiD estimate tells us whether employment changed more in NJ than in PA after the policy.
If positive: employment rose more in NJ
If near zero: employment did not fall relative to PA
If negative: employment fell more in NJ
## Wages
The wage DiD estimate tells us whether wages changed more in NJ than in PA.
If positive: wages rose more in NJ
This is what we expect if the policy affected wages

## Critical Question:
Does a positive DiD coefficient automatically prove causality?

Not automatically. It supports a causal interpretation only if the DiD assumptions are credible, especially parallel trends and the absence of major confounding shocks that affected the treated group differently.
