---
title: "Lab 8: Multilevel Models: Students Within Schools"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---
# Overview
In previous labs, we assumed that all observations were independent. However, in many real datasets this assumption is unrealistic. Observations are often grouped into higher-level units. For example:

• Patients are treated within hospitals  
• Students are taught within schools  
• Employees work within firms  
• Residents live within counties  

When observations belong to the same group, they often share common influences. For example, students in the same school share:

• teachers  
• curriculum  
• administrative policies  
• peer environments  
• resource availability  

Because of this, their outcomes may be correlated. Standard regression assumes independence of observations. When this assumption is violated, standard errors may be incorrect and inference may be misleading.

Multilevel models (also called hierarchical models or mixed effects models) explicitly model this structure.

In this lab we will analyze student math achievement where students are nested within schools.

We will:

• Compare OLS vs multilevel models  
• Measure between-school variation  
• Compute the Intraclass Correlation (ICC)  
• Fit random intercept models  
• Fit random slope models  
• Interpret variance components  

---

## Learning Objectives

By the end of this lab you should be able to:

• Explain why clustering violates regression assumptions  
• Interpret random intercept models  
• Calculate ICC  
• Explain variance decomposition  
• Interpret school-level variation  
• Understand partial pooling  

---

## Load data
This dataset contains student level and school level variables merged into one dataset.
Student variables:
• MathAch → math achievement score
• SES → socioeconomic status
• Female → gender indicator
• Minority → minority indicator
School variables:
• Sector → public vs Catholic school
• Size → school size
• MeanSES → average school SES
• Disclim → discipline climate

```{r}
# Set CRAN mirror (REQUIRED for Quarto rendering)
options(repos = c(CRAN = "https://cloud.r-project.org"))

required_packages <- c("tidyverse","lme4","lmerTest")

new_packages <- required_packages[
  !(required_packages %in% installed.packages()[,"Package"])
]

if(length(new_packages)){
  install.packages(new_packages)
}

lapply(required_packages, library, character.only = TRUE)

hsb <- read.csv("hsb_multilevel.csv")

hsb$Sex <- factor(hsb$Sex)
hsb$Minority <- factor(hsb$Minority)
hsb$Sector <- factor(hsb$Sector)
hsb$School <- factor(hsb$School)

# Remove duplicate column
hsb$MEANSES.y <- NULL

str(hsb)
head(hsb)
```

---

# Data exploration
## Understanding the hierarchical structure
Before fitting models, we should understand the structure of the data.
```{r}
summary(hsb)

table(hsb$School)

length(unique(hsb$School))

nrow(hsb)

```

Some questions to consider:
- How many schools are in the dataset?
- How many students are in each school?
- What is the average math achievement score across all students?
- What is the average math achievement score within each school?
- How much variation in math achievement is there between schools vs within schools?
- Are there any schools that stand out as particularly high or low performing?
- What is the average number of students per school?
- Why might students within the same school have similar outcomes?
- Give three possible shared influences within schools.

# Exploring Math Achievement
```{r}
summary(hsb$MathAch)

ggplot(hsb,aes(MathAch))+
geom_histogram(bins=30)+
theme_minimal()
```

Some questions to consider:
- Is math achievement approximately normally distributed?
- Are there any extreme values?
- Why might distribution shape matter for modeling?

# Baseline OLS model
First, let's fit a simple OLS regression ignoring the hierarchical structure.
This assumes:
All students are independent.
This is likely false because students share schools.
```{r}
ols <- lm(MathAch ~ SES + Sex + Minority, data=hsb)

summary(ols)
```

Some interpretation questions:
- Interpret the SES coefficient.
- Does being female significantly affect math achievement?
- Does being a minority significantly affect math achievement?
- What assumptions does OLS make about errors?
- Why might those assumptions fail here?

# Why OLS may be wrong here
OLS assumes:
$Cov(\epsilon_i,\epsilon_j)=0$
But if students share schools:
$Cov(\epsilon_i,\epsilon_j) \neq 0$
This creates correlated residuals.
When this happens:
• Standard errors are underestimated
• Significance may be overstated
• Inference becomes unreliable
Multilevel models fix this by modeling school variation explicitly.

# Null Multilevel Model
Let's fit a null model with no predictors, just random intercepts for schools.
This partitions variance into:
- Between-school variance
- Within-school variance
```{r}
null_model <- lmer(MathAch ~ 1 + (1 | School), data = hsb)

summary(null_model)
```

# Understanding Variance Components
This model estimates:
MathAch = school effect + student noise
Variance components represent:
- School variance → differences between schools
- Residual variance → differences within schools

# ICC Calculation
The Intraclass Correlation Coefficient (ICC) measures the proportion of total variance that is between schools.
$ICC = \frac{\sigma^2_{school}}{\sigma^2_{school}+\sigma^2_{residual}}$
A higher ICC means more clustering within schools. 
```{r}
var_school <- as.numeric(VarCorr(null_model)$School)

var_resid <- attr(VarCorr(null_model),"sc")^2

ICC <- var_school/(var_school+var_resid)

ICC
```

Some interpretation questions:
- What does the ICC value tell us about the clustering of math achievement within schools?
- If the ICC is close to 0, what does that imply about the importance of school-level factors?
- If the ICC is close to 1, what does that imply about the importance of school-level factors?
- How would ignoring the ICC affect our conclusions about the effects of SES, gender, and minority status on math achievement?
- What are some potential school-level factors that could explain between-school variation in math achievement?
- How might we include those school-level factors in a multilevel model?
- What are the benefits of using a multilevel model compared to OLS in this context?
- Why does ICC justify the use of multilevel modeling for this data?

# Random Intercept Model
Now let's add student-level predictors to the model while still allowing for random intercepts by school.
```{r}
model1 <- lmer(MathAch ~ SES + Sex + Minority + (1 | School),
               data = hsb)
summary(model1)
```

Some interpretation questions:
- How do the coefficients for SES, Female, and Minority change compared to the OLS model?
- Are the predictors still significant after accounting for school-level variation?
- What does the random intercept variance tell us about differences between schools after controlling for student-level factors
- How would you interpret the random intercepts for each school?
- What does it mean if a school has a positive random intercept vs a negative random intercept?
- How does the inclusion of random intercepts affect our understanding of the effects of SES, gender, and minority status on math achievement?
- What are some potential limitations of this random intercept model?
- How might we extend this model to include school-level predictors?

# Adding School-Level Predictors
Now let's add school-level predictors to explain some of the between-school variation.
```{r}
model2 <- lmer(MathAch ~ SES + Sex + Minority + Sector + MEANSES.x + (1 | School),
               data = hsb)
summary(model2)

```

Some interpretation questions:
- How do the coefficients for Sector and MeanSES help explain between-school variation in math achievement?
- Do these school-level predictors reduce the random intercept variance compared to the previous model?
- How do the student-level predictors change after adding school-level predictors?
- What does this tell us about the relative importance of student-level vs school-level factors in explaining math achievement?
- How would you interpret the coefficient for Sector (public vs Catholic schools)?
- How would you interpret the coefficient for MeanSES (average school SES)?
- What are some potential confounders that could affect the relationship between school-level predictors and math achievement?
- How might we further extend this model to include interactions between student-level and school-level predictors?

# Random Slope Model
Finally, let's allow the effect of SES to vary across schools by adding a random slope for SES. This tests whether SES matters more in some schools.
```{r}
model3 <- lmer(MathAch ~ SES + Sex + Minority + (SES | School),
               data = hsb)
summary(model3)

```

Some interpretation questions:
- How does the random slope for SES affect our understanding of the relationship between SES and math achievement?
- What does the variance of the random slope for SES tell us about how much the effect of SES varies across schools?
- Are there any schools where SES has a particularly strong or weak effect on math achievement?
- How does the inclusion of a random slope for SES affect the fixed effect coefficient for SES?
- What does the correlation between the random intercept and random slope for SES tell us about the relationship between school-level effects and the effect of SES?
- How would you interpret a positive vs negative correlation between the random intercept and random slope for SES?
- What are some potential implications of finding significant random slopes for SES in terms of educational policy and interventions?

# Model Comparison
We can compare the fit of our models using AIC or likelihood ratio tests.
```{r}
AIC(ols,model1,model2,model3)
```

Some questions to consider:
- Which model has the lowest AIC and is therefore the best fitting model?
- How does the inclusion of random effects and school-level predictors improve model fit compared to OLS?
- What does this tell us about the importance of accounting for hierarchical structure in the data?
- What is the tradeoff between model complexity and fit?

# Visualizing School Effects
We can visualize the random intercepts for each school to see which schools are performing above or below average after controlling for student-level factors. 

```{r}
ranef(model1)$School %>%
ggplot(aes(x=`(Intercept)`))+
geom_histogram(bins=30)+
theme_minimal()
```

Some questions to consider:
- What does the distribution of random intercepts tell us about the variation in school effects on math achievement?
- Are there any schools that stand out as particularly high or low performing after controlling for student-level factors?
- How might we use this information to target interventions or resources to schools that are struggling?
- What are some potential limitations of interpreting random intercepts as measures of school quality?

# Key Takeaways
Multilevel models allow us to:
• Model clustered data
• Estimate group variation
• Improve inference
• Separate within and between effects
They are essential whenever data has hierarchical structure.